Manuscript for Foodborne-Outbreak Project

### 2023-04-21 code/analysis_code/exploratory_analysis.qmd
update

## Testing/Analysis Data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_analysis_data.rds")
# Import Data
analysis_data_final <- readRDS(data_location)
```

@@

```{r fig.height=8,fig.width=8}
# Time vs Hospitalizations
ggplot(training_data_final %>% 
         group_by(Year, Pathogen_Type, Hospitalizations) %>% 
         count() %>% 
         mutate(true_hospitalization = n * Hospitalizations), 
       aes(x = Year, y = true_hospitalization, 
           group = Pathogen_Type, color = Pathogen_Type)) + 
  geom_point() + geom_jitter() + 
  labs(x = "Year", y = "Hospitalizations", 
       title = "Scatter Plot of Hospitalizations by Year and Pathogen Type")

## Save Graph
image_path <- here("results","images_EDA")
ggsave("time_v_hospitalization_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")

# Pathogen Type + Etiology vs Hospitalization
ggplot(training_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% count() %>% 
         mutate(true_hospitalization = n * Hospitalizations), 
       aes(x = Pathogen_Type, y = true_hospitalization, 
           group = Simplified_Etiology, fill = Simplified_Etiology)) +
  geom_col() + 
  labs(x = "Pathogen Type", y = "Hospitalizations", 
       title = "Stacked Bar Chart of Hospitalizations by Pathogen Type and Etiology")

## Save Graph
ggsave("pathogen_type_etiology_v_hospitalizations.png", path = image_path, width = 3000, height = 2000, units = "px")

## Closer Look at Bacterial Etiologies for Outbreaks
ggplot(training_data_final %>% 
         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + 
  labs(x = "Bacteria", y = "Hospitalizations", 
       title = "Bar Chart of Bacteria Frequency") +
  coord_flip()

@@

### Even Closer Look at Salmonella Cases

Salmonella_Hosp <- training_data_final %>%
  select(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
  group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
  filter(Simplified_Etiology == "Salmonella") %>%
  filter(`IFSAC Category` != "Multiple") %>%
  filter(`IFSAC Category` != "Other") %>%
  count() %>%
  mutate(total_hospitalizations = n * Hospitalizations) %>%
  arrange(desc(total_hospitalizations))

ggplot(Salmonella_Hosp, aes(x=`IFSAC Category`, y = total_hospitalizations))+
  geom_boxplot() +
  coord_flip()

# States vs Hospitalization

state_v_hospitalization_training <- training_data_final %>% 
  group_by(State) %>% 
  mutate(all_hospitalizations = sum(Hospitalizations)) %>% 
  count(all_hospitalizations) %>% 
  select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))

@@

## Project Total Hospitalizations onto State Map
training_hosp <- training_data_final%>%
  mutate(state = tolower(State))%>%
  group_by(state)%>%
  mutate(all_hosp = sum(Hospitalizations)) %>% count(all_hosp) %>%
  select(state, all_hosp)
training_hosp

@@

map_final %>%
  ggplot(aes(x=long, y=lat, 
             fill = all_hosp, group = group)) +
  geom_polygon(color="black", size = 0.1) +
  geom_polygon(data = state_map, aes(x= long, y= lat, group = group), 
               fill = NA, color = "black", size= 0.3) +
  scale_fill_gradient(name = "Number of \n Hospitalizations",
                      low = "#FFFFFF", high="#FF0000",
                       limits = c(0,NA)) +
  labs(title = "Cases of Hospitalizations Caused by Foodborne Illnesses\n in Each State During 1998-2015") +
  theme_void() +
  coord_quickmap()

ggsave("State_hospitalization_map.png", path = image_path, width=3000, height=2000, units="px")

# Food Type and Hospitalizations
ggplot(training_data_final %>% 
         filter(`IFSAC Category` != "Multiple") %>% 
         filter(`IFSAC Category` != "Other"), 
       aes(x = `IFSAC Category`, y = Hospitalizations)) + 
  geom_col() + 
  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
  coord_flip()

@@

# Locations and Hospitalizations
Location_v_Hospitalization <- ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
  geom_col() + 
  labs(x = "Locations", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
  coord_flip()

@@

## Testing Data
```{r}
# Illnesses Caused by Pathogen Type Over Time
ggplot(analysis_data_final %>% 
         group_by(Year, Pathogen_Type, Illnesses) %>% 
         count() %>% 
         mutate(true_illness_count = n * Illnesses), 
       aes(x = Year, y = true_illness_count, group = Pathogen_Type, color = Pathogen_Type)) + 
  geom_point() + 
  geom_jitter() + 
  labs(x = "Year", y = "Cases", title = "Scatter Plot of Illness by Year and Pathogen Type")

## Save Graph
ggsave("time_v_illness_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")

# Pathogen Type + Etiology vs Illnesses Caused
ggplot(analysis_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
         count() %>% 
         mutate(true_illness_count = n * Illnesses), 
       aes(x = Pathogen_Type, y = Illnesses, group = Simplified_Etiology, fill = Simplified_Etiology)) + 
  geom_col() + 
  labs(x = "Pathogen Type", y = "Cases", title = "Stacked Bar Chart of Illnesses by Pathogen Type and Etiology")

## Save Graph
ggsave("pathogen_type_etiology_v_illness.png", path = image_path, width = 3000, height = 2000, units = "px")

## Closer Look at Bacterial Etiologies
ggplot(analysis_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
         filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + 
  geom_bar() + 
  labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")+
  coord_flip()

# Save Graph
ggsave("bacterial_etiology_v_outbreak_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")

# States vs Illnesses

state_v_illness_analysis <- analysis_data_final %>% 
  group_by(State) %>%
  mutate(all_illnesses = sum(Illnesses)) %>%
  count(all_illnesses) %>% 
  select(State, all_illnesses) %>% 
  arrange(desc(all_illnesses))

saveRDS(state_v_illness_analysis, file = table_path)

rmarkdown::paged_table(state_v_illness_analysis)

# Food Type and Illnesses
ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + 
  geom_col() + 
  labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
  coord_flip()

# Save Graph
ggsave("food_type_v_illness_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")

# Locations and Illnesses
ggplot(analysis_data_final, aes(x = Location, y = Illnesses)) +
  geom_col() + 
  labs(x = "Locations", y = "Illnesses", title = "Illnesses by Locations")+
  coord_flip()

# Save Graph
ggsave("Location_v_Illness.png", path = image_path, width = 3000, height = 2000, units = "px")

```
The outbreaks with a high number of cases (> 500) are caused by bacteria. The stacked barplot indicates that the top two etiologic agents that contribute to fooborne-illnesses are Salmonella and Norovirus, which belong to the pathogen type of Bacteria and Virus respectively. In bacterial pathogen, Salmonella genus is the most dominant factor that is associated with foodborne outbreaks, followed by E. coli species and Vibrio genus. Geographically speaking, Arkansas and Minnesota are the two states with the highest accumulated foodborne illnesses from 1998 to 2015. Salad, Poultry, and Fruits are the three most common types of food vectors that are involved in transmitting foodborne pathogens. Resturants and Private home are the most frequent locations that foodborne outbreaks occur.
Overall, the relationships between predictors and outcomes of training dataset share similar patterns with those of analysis dataset.

>>>>

## Testing/Analysis Data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_testing_data.rds")
# Import Data
analysis_data_final <- readRDS(data_location)
```

@@

### Pathogen Type + Etiology vs Hospitalization
```{r fig.height=8,fig.width=8}
# Pathogen Type + Etiology vs Hospitalization
ggplot(training_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% 
         count() %>% mutate(true_hospitalization = n * Hospitalizations), 
       aes(x = Pathogen_Type, y = true_hospitalization, 
           group = Simplified_Etiology, fill = Simplified_Etiology)) +
  geom_col() + 
  labs(x = "Pathogen Type", y = "Hospitalizations", 
       title = "Stacked Bar Chart of Hospitalizations by Pathogen & Etiology") +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold", size = 11, hjust = 0.5))

## Save Graph
image_path <- here("results","images_EDA")
ggsave("pathogen_type_etiology_v_hospitalizations.png", path = image_path, width = 3000, height = 2000, units = "px")
```

### Template Setup
```{r}
# Template Graph
theme_set(theme_bw())
theme_update(plot.title = element_text(face = "bold", size = 12))
theme_update(plot.title = element_text(hjust = 0.5))
```

### Time vs Hospitalizations
```{r}
# Data Visualization
## Illnesses
# ggplot(training_data_final %>%
#          group_by(Year, Pathogen_Type, Illnesses) %>%
#          count() %>%
#          mutate(true_illnesses = n * Illnesses),
#        aes(x = Year, y = true_illnesses,
#            group = Pathogen_Type, color = Pathogen_Type)) +
#   geom_point() + geom_jitter() +
#   labs(x = "Year", y = "Illnesses",
#        title = "Scatter Plot of Illnesses by Year and Pathogen Type",
#        color = "Pathogen Type") +
#   facet_wrap(~ Pathogen_Type) +
#   theme(legend.position = "none")

## Hospitalizations
ggplot(training_data_final %>% 
         group_by(Year, Pathogen_Type, Hospitalizations) %>% 
         count() %>% 
         mutate(true_hospitalization = n * Hospitalizations), 
       aes(x = Year, y = true_hospitalization, 
           group = Pathogen_Type, color = Pathogen_Type)) + 
  geom_point() + geom_jitter() + 
  labs(x = "Year", y = "Hospitalizations", 
       title = "Scatter Plot of Hospitalizations by Year and Pathogen Type",
       color = "Pathogen Type") + 
  facet_wrap(~ Pathogen_Type) + 
  theme(legend.position = "none")

## Save Graph
ggsave("time_v_hospitalization_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")
```

### Bacterial Etiologies for Outbreaks
```{r}
# General
ggplot(training_data_final %>% 
         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>%
         filter(Pathogen_Type == "Bacteria"), 
       aes(x = Simplified_Etiology)) + geom_bar() + 
  labs(x = "Bacteria", y = "Hospitalizations", 
       title = "Bar Chart of Bacteria Frequency") + coord_flip()

@@

# Salmonella
Salmonella_Hosp <- training_data_final %>%
  select(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
  group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
  filter(Simplified_Etiology == "Salmonella") %>%
  filter(`IFSAC Category` != "Multiple") %>%
  filter(`IFSAC Category` != "Other") %>%
  count() %>%
  mutate(total_hospitalizations = n * Hospitalizations) %>%
  arrange(desc(total_hospitalizations))

ggplot(Salmonella_Hosp, aes(x=`IFSAC Category`, y = total_hospitalizations))+
  geom_boxplot() + labs(x = "Total Hospitalizations", y = "Food Type", 
                        title = "Hospitalizations by Food Type for Salmonella") +
  coord_flip()
```
### Hospitalizations by State
```{r}
# States vs Hospitalization
state_v_hospitalization_training <- training_data_final %>% 
  group_by(State) %>% 
  mutate(all_hospitalizations = sum(Hospitalizations)) %>% 
  count(all_hospitalizations) %>% 
  select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))

training_data_final %>% 
  group_by(State) %>% 
  mutate(all_illnesses = sum(Illnesses)) %>% 
  count(all_illnesses) %>% 
  select(State, all_illnesses) %>% arrange(desc(all_illnesses))

@@

### State Map of Hospitalizations
```{r}
training_hosp <- training_data_final%>%
  mutate(state = tolower(State))%>%
  group_by(state)%>%
  mutate(all_hosp = sum(Hospitalizations)) %>% count(all_hosp) %>%
  select(state, all_hosp)
training_hosp

@@

map_final %>%
  ggplot(aes(x=long, y=lat, 
             fill = all_hosp, group = group)) +
  geom_polygon(color="black", size = 0.1) +
  geom_polygon(data = state_map, aes(x= long, y= lat, group = group), 
               fill = NA, color = "black", size= 0.3) +
  scale_fill_gradient(name = "Number of \n Hospitalizations",
                      low = "#FFFFFF", high="#FF0000",
                       limits = c(0,NA)) +
  labs(title = "Cases of Hospitalizations Caused by Foodborne Illnesses in Each State (1998-2015)") +
  theme_void() + theme(plot.title = element_text(hjust = 0.5)) + 
  theme(plot.title = element_text(face = "bold", size = 12)) +
  coord_quickmap()

ggsave("State_hospitalization_map.png", path = image_path, width = 3000, height = 2000, units="px")
```

### Food Type and Hospitalizations
```{r}
# Food Type and Hospitalizations
ggplot(training_data_final %>% 
         filter(`IFSAC Category` != "Multiple") %>% 
         filter(`IFSAC Category` != "Other") %>%
         group_by(`IFSAC Category`) %>% 
         mutate(all_hospitalizations = sum(Hospitalizations)) %>%
         count(all_hospitalizations) %>% arrange(desc(all_hospitalizations)), 
       aes(x = `IFSAC Category`, y = all_hospitalizations, 
           fill = `IFSAC Category`)) + 
  geom_col() + 
  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'") +
  coord_flip() + theme(legend.position = "none")

@@

# Extra: Food Type & Illnesses
training_data_food_and_illnesses <- training_data_final %>% 
         filter(`IFSAC Category` != "Multiple") %>% 
         filter(`IFSAC Category` != "Other") %>%
         group_by(`IFSAC Category`) %>% 
         mutate(all_illnesses = sum(Illnesses)) %>%
         count(all_illnesses) %>% 
  arrange(desc(all_illnesses))
table_path <- here("results", "training_food_vs_illnesses.rds")
saveRDS(training_data_food_and_illnesses, file = table_path)

### Locations and Hospitalizations
```{r fig.height=8,fig.width=8}
Location_v_Hospitalization <- ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
  geom_col() + 
  labs(x = "Locations", y = "Hospitalizations", title = "Outbreak Location versus Hospitalizations") +
  coord_flip()

@@

## Testing Data

### Pathogen Type vs Illnesses Over Time
```{r}
ggplot(analysis_data_final %>% 
         group_by(Year, Pathogen_Type, Illnesses) %>% 
         count() %>% 
         mutate(true_illness_count = n * Illnesses), 
       aes(x = Year, y = true_illness_count, group = Pathogen_Type, color = Pathogen_Type)) + 
  geom_point() + 
  geom_jitter() + 
  labs(x = "Year", y = "Cases", title = "Scatter Plot of Illness by Year and Pathogen Type")

ggsave("time_v_illness_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")

### Pathogen Type and Etiology vs Illnesses
```{r}
# Visualization
ggplot(analysis_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
         count() %>% 
         mutate(true_illness_count = n * Illnesses), 
       aes(x = Pathogen_Type, y = Illnesses, group = Simplified_Etiology, fill = Simplified_Etiology)) + 
  geom_col() + 
  labs(x = "Pathogen Type", y = "Cases", title = "Stacked Bar Chart of Illnesses by Pathogen Type and Etiology")

# Save Graph
ggsave("pathogen_type_etiology_v_illness.png", path = image_path, width = 3000, height = 2000, units = "px")
```

### Bacterial Etiologies
```{r}
ggplot(analysis_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
         filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + 
  geom_bar() + 
  labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency for Illnesses")+
  coord_flip()

# Save Graph
ggsave("bacterial_etiology_v_outbreak_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
```

### Illnesses by State
```{r}
state_v_illness_analysis <- analysis_data_final %>% 
  group_by(State) %>%
  mutate(all_illnesses = sum(Illnesses)) %>%
  count(all_illnesses) %>% 
  select(State, all_illnesses) %>% 
  arrange(desc(all_illnesses))

# Save Table
table_path <- here("results", "state_v_illness_training.rds")
saveRDS(state_v_illness_analysis, file = table_path)

rmarkdown::paged_table(state_v_illness_analysis)
```

### Food Type and Illnesses
```{r}
ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + 
  geom_col() + 
  labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
  coord_flip()

ggsave("food_type_v_illness_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")

### Locations and Illnesses
```{r}
# Locations and Illnesses
ggplot(analysis_data_final %>% group_by(Location) %>% 
         mutate(all_illnesses = sum(Illnesses)) %>%
         count(all_illnesses) %>% arrange(desc(all_illnesses)), aes(x = Location, y = all_illnesses)) +
  geom_col() + 
  labs(x = "Locations", y = "Illnesses", title = "Illnesses by Location")+
  coord_flip()

# Save Graph
ggsave("Location_v_Illness.png", path = image_path, width = 3000, height = 2000, units = "px")
```
The outbreaks with a high number of cases (> 500) are caused by bacteria. The stacked barplot indicates that the top two etiologic agents that contribute to fooborne-illnesses are Salmonella and Norovirus, which belong to the pathogen type of Bacteria and Virus respectively. In bacterial pathogen, Salmonella genus is the most dominant factor that is associated with foodborne outbreaks, followed by E. coli species and Vibrio genus. Geographically speaking, Arkansas and Minnesota are the two states with the highest accumulated foodborne illnesses from 1998 to 2015. Salad, Poultry, and Fruits are the three most common types of food vectors that are involved in transmitting foodborne pathogens. Resturants and Private home are the most frequent locations that foodborne outbreaks occur.

Overall, the relationships between predictors and outcomes of training dataset share similar patterns with those of analysis dataset.

### 2023-04-21 code/processing_code/processingfile_v1.qmd
update

# Match Training Data Format
testing_data_fin <- testing_data_fin %>%
  mutate(Location = case_when(
    grepl("Private Home/Residence", Location, ignore.case = TRUE) ~ "Private Home/Residence",
    grepl("Caterer", Location, ignore.case = TRUE) ~ "Catering Service",
    grepl("Office/Indoor Workspace", Location, ignore.case = TRUE) ~ "Office/Indoor Workspace",
    grepl("Banquet", Location, ignore.case = TRUE) ~ "Banquet Facility",
    grepl("Fair", Location, ignore.case = TRUE) ~ "Fair/Festival",
    grepl("Religious", Location, ignore.case = TRUE) ~ "Religious Facility",
    grepl("Grocery Store", Location, ignore.case = TRUE) ~ "Grocery Store",
    grepl("School/College/University", Location, ignore.case = TRUE) ~ "School/College/University",
    grepl("Prison/Jail", Location, ignore.case = TRUE) ~ "Prison/Jail",
    grepl("Long-Term Care", Location, ignore.case = TRUE) ~ "Nursing Home/Assisted Living Facility",
    grepl("Child", Location, ignore.case = TRUE) ~ "Child Daycare",
    grepl("Farm/Dairy", Location, ignore.case = TRUE) ~ "Farm/Dairy",
    grepl("Hotel/Motel", Location, ignore.case = TRUE) ~ "Hotel/Motel",
    TRUE ~ Location))
```

>>>>

# Match Training Data Format
testing_data_fin <- testing_data_fin %>%
  mutate(Location = case_when(
    grepl("Private Home/Residence", Location, ignore.case = TRUE) ~ "Private Home/Residence",
    grepl("Caterer", Location, ignore.case = TRUE) ~ "Catering Service",
    grepl("Office/Indoor", Location, ignore.case = TRUE) ~ "Office/Indoor Workplace",
    grepl("Banquet", Location, ignore.case = TRUE) ~ "Banquet Facility",
    grepl("Fair", Location, ignore.case = TRUE) ~ "Fair/Festival",
    grepl("Religious", Location, ignore.case = TRUE) ~ "Religious Facility",
    grepl("Grocery Store", Location, ignore.case = TRUE) ~ "Grocery Store",
    grepl("School/College/University", Location, ignore.case = TRUE) ~ "School/College/University",
    grepl("Prison/Jail", Location, ignore.case = TRUE) ~ "Prison/Jail",
    grepl("Long-Term Care", Location, ignore.case = TRUE) ~ "Nursing Home/Assisted Living Facility",
    grepl("Child", Location, ignore.case = TRUE) ~ "Child Daycare",
    grepl("Farm/Dairy", Location, ignore.case = TRUE) ~ "Farm/Dairy",
    grepl("Hotel/Motel", Location, ignore.case = TRUE) ~ "Hotel/Motel",
    grepl("Ship/Boat", Location, ignore.case = TRUE) ~ "Ship/Boat",
    TRUE ~ Location))

unique(testing_data_fin$Location)
unique(training_data_fin$Location)
```

### 2023-04-21 products/manuscript/Manuscript2.qmd
update

---
title: "Manuscript for Foodborne-Outbreak Project"
author: "Weifan Wu & Kailin (Kai) Chen"
date: "May 2023"
format:
  docx
bibliography: ../references.bib
csl: ../apa.csl
---

```{r echo = FALSE, warning = FALSE, paged.print = TRUE}
# Load Libraries
library(here)
library(knitr)
library(tidyverse)
library(broom)
library(ggtext)
```

@@

Foodborne pathogens typically cause serious gastrointestinal infections. On average, 1 in 6 Americans will acquire foodborne illness each year; 128,000 will be hospitalized, and 3,000 will die [@centerforfoodsafetyandappliednutrition2022]. Economic loss associated with foodborne illness is estimated to be greater than 50 billion US dollars annually. Restaurants are frequently sites attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020; @sarno2021]. Disease etiologies range from bacteria (ex: *Salmonella* spp. and *E. coli*) to viruses (norovirus and rotavirus), to parasites (*Cryptosporidium* spp. and *Cyclospora* spp. [@cdc2022]. To verify pathogens of greatest concern, common settings associated with foodborne illness, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).

@@

From this research project, we would like to answer the following questions: - What are the most frequent locations that foodborne outbreaks occur? - What pathogen species are most frequently associated with foodborne outbreaks? - What type of food is most often seen for foodborne outbreaks?

@@

Our hypotheses include: - Outbreaks occur most often in homes rather than restaurants. - The most common pathogen associated with foodborne disease outbreaks is *E. coli*. - The food most commonly associated with foodborne outbreaks consists of salad.

@@

Two datasets were adopted for this research. The first dataset--which was used to build statistical models for the study--was uploaded onto *Kaggle* in 2017, with 12 variables and more than 16,950 observations [@cdc2017a]. The second dataset--which was used for machine learning--was retrieved from the CDC's National Outbreak Reporting System (NORS) website. This second dataset covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.

@@

![Figure 2: Hospitalizations by Food Type.](images/Food_type_v_hospitalization-01.png)

```{r Visualization of Hospitalizations by State}
rmarkdown::paged_table(readRDS("../../results/state_v_hospitalization_training.rds"))
```

@@

![Figure 1. Year vs Illnesses Caused by Pathogen Type](images/time_v_illness_pathogen_type.png)

![Figure 3. Illnesses by outbreak location.](images/Location_v_Illness.png)

![Figure 4. Bacterial etiology by outbreak frequency.](images/bacterial_etiology_v_outbreak_frequency.png)

@@

@@

>>>>

---
title: "Manuscript for Foodborne-Outbreak Project"
author: "Weifan Wu & Kailin (Kai) Chen"
date: "May 2023"
format:
  html
bibliography: ../references.bib
csl: ../apa.csl
---

```{r echo = FALSE, warning = FALSE, paged.print = TRUE}
# Load Libraries
library(here)
library(knitr)
library(tidyverse)
library(broom)
library(ggtext)
library(flextable)
```

@@

Foodborne pathogens can cause serious gastrointestinal infections. About 600 million cases of foodborne disease are reported worldwide, with almost 1/10 of these cases caused by contaminated food [@he2023]. On average within the United States each year, 1 in 6 Americans will acquire foodborne illness; 128,000 will be hospitalized, and 3,000 will die [@centerforfoodsafetyandappliednutrition2022]. Economic loss associated with food poisoning is estimated to be greater than 50 billion US dollars annually. Restaurants are frequently sites attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020; @sarno2021]. Disease etiologies range from bacteria (ex: *Salmonella* spp. and *E. coli*) to viruses (norovirus and rotavirus), to parasites (*Cryptosporidium* spp. and *Cyclospora* spp. [@cdc2022]. Environmental factors such as heavy metals can also cause foodborne illness.

@@

To verify pathogens of greatest concern, common settings associated with food poisoning, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).

From this research project, we would like to answer the following questions:

-   What are the most frequent locations that foodborne outbreaks occur?

-   What pathogen species are most frequently associated with foodborne outbreaks?

-   What type of food is most often seen for foodborne outbreaks?

@@

Our hypotheses include:

-   Outbreaks occur most often in homes rather than restaurants.

-   The most common pathogen associated with foodborne disease outbreaks is *E. coli*.

-   The food most commonly associated with foodborne outbreaks consists of salad.

@@

Two datasets were adopted for this research. The first dataset--which was used to build statistical models for the study--was uploaded onto *Kaggle* in 2017, with 12 variables and more than 16,950 observations spanning almost two decades from 1998 to 2015 [@cdc2017a]. The second dataset--which was used for machine learning--was retrieved from the CDC's National Outbreak Reporting System (NORS) website. This second dataset covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.

@@

```{r Table of Hospitalizations by State}
first_table <- readRDS("../../results/state_v_hospitalization_training.rds") %>% 
  rename(Hospitalizations = all_hospitalizations)
first_table <- flextable(first_table)
first_table <- theme_vanilla(first_table)
first_table <- set_caption(first_table, caption = as_paragraph(colorize("Table 1. Hospitalizations by state from 1998 to 2015.", color = "black")))
first_table <- set_table_properties(first_table, layout = "autofit")
first_table
```

![Figure 1. Hospitalizations by food type from 1998-2015.](images/Food_type_v_hospitalization-01.png)

```{r}
second_table <- readRDS("../../results/training_food_vs_illnesses.rds") %>% 
  select(`IFSAC Category`, all_illnesses) %>% 
  rename(Illnesses = all_illnesses)
second_table <- flextable(second_table)
second_table <- theme_vanilla(second_table)
second_table <- set_caption(second_table, caption = as_paragraph(colorize("Table 2. Illnesses by food type from 1998 to 2015.", color = "black")))
second_table
```

@@

![Figure 2. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png)

![Figure 3. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness-01.png)

![Figure 4. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png)

@@

Exploratory data visualization demonstrates bacteria contribute to the majority outbreaks, and all outbreaks with illnesses numbering over 1,000 and hospitalizations over 100 are attributed to bacteria. From all pathogen types, bacteria are thus associated with worse health outcomes when it comes to food poisoning.

From 1998 to 2015, multistate outbreaks caused the most illnesses (N = 18,617) and hospitalizations (3,506) than any in-state outbreak. The fact that multistate outbreaks were seen more often than in any individual state for illnesses and hospitalizations suggests that these disease outbreaks typically involve a regional or national ingredient distributor. Hence, it would be reasonable to suggest laboratory sampling of goods, especially for bacteria, as a safety benchmark upon arrival at their destinations.

In terms of food categories, poultry was identified to have caused the most hospitalizations (N = 770), followed closely by fruits and beef. Interestingly, however, salads were found to cause the most illnesses (N = 11,493). Upon addition of total hospitalizations and illnesses for both poultry and salads, salads were determined to have the highest number of cases in total, which confirms our original hypothesis.

The remaining two of the three initial hypotheses were proved incorrect. Instead of *E. coli*, results overall indicated that *Salmonella* spp. are the most common pathogens linked to foodborne disease; these bacteria contribute to the greatest amount of illnesses and hospitalizations in comparison to other disease agents for our data. Finally, we also see that food poisoning is most frequently linked to restaurants instead of private households.

\[Something about machine learning here???\]

These results offer insight into common disease characteristics across food poisoning cases, which is greatly useful for epidemiologists and other public health professionals to anticipate which areas to investigate first for future outbreaks.

@@

Food poisoning in the United States poses a great concern to public health in terms of both morbidity and economic costs. Previous surveillance data points to *Salmonella* bacteria and salads as the main cause and medium for foodborne disease outbreaks, along with the most common setting for such outbreaks being a restaurant. These findings will also give future outbreak investigators a first direction to consider in the field.
